Forecasting the 2024 US Presidential Election

### 2024-10-22 paper/paper.qmd
added data section

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

@@

## Overview
We obtained the "Presidential general election polls" dataset from FiveThirtyEight [@FiveThirtyEight2024] and used the statistical programming language R [@citeR] to analyze this data.

*Person A:* Provide an overview of the dataset's context, sources, and key variables.

## Measurement
Describe how the real-world phenomenon (voter preference) is translated into a dataset entry.

*Person A:* Discuss the survey design, the process of data collection, and potential biases or errors in measurement.

## Outcome variables
Add graphs, tables, and text for each outcome variable. Use subsections as needed.

Some of our data is (@fig-outcome), from @source

```{r}
#| label: fig-outcome
#| fig-cap: Example Outcome Variable Visualization
#| echo: false
# Visualization code

@@

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables
Add graphs, tables, and text to describe key predictors.

*Person A:*: Visualize variable distributions (e.g., histograms, boxplots), and justify the inclusion of predictors in the model.

```{r}
#| label: fig-predictors
#| fig-cap: Example Predictor Variable Visualization
#| echo: false
# Visualization code

@@

# Pollster Methodology Overview (Appendix A)
Append a detailed review of a chosen pollster's methodology, survey techniques, strengths, and weaknesses.

@@

# Results

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
library(tidyverse)
library(dplyr)
library(knitr)
library(lubridate)
library(ggplot2)
library(ggcorrplot)
library(kableExtra)

@@

## Overview
We sourced the "Presidential General Election Polls" dataset from FiveThirtyEight [@FiveThirtyEight2024] and performed an in-depth analysis using the statistical programming language R [@citeR].

Our primary objective was to clean, organize, and analyze U.S. presidential election polling data to provide insights into voter preferences by state. This involved selecting data from reputable pollsters, organizing it by state, and addressing any missing or incomplete entries. Additional datasets were merged where necessary, and after thoroughly addressing missing values, we finalized a polished dataset ready for analysis. Below is a table summarizing the polling data by state, showcasing voter preferences by party and demographic breakdowns. The table highlights key trends and differences across states in terms of party lean and demographic distribution. @fig-predictors-pct-summary provides a snapshot of these findings.

```{r}
#| echo: false
#| warning: false
#| message: false

set.seed(754)
poll_data <- read_csv(here::here(
  "data/02-analysis_data/organized_state_poll_data.csv"))

# Name references
column_refs <- list(
  democrat = "democrat_pct",
  republican = "republican_pct",
  highly_educated = "What is the highest level of education you have attained? College graduate",
  lower_educated = "What is the highest level of education you have attained? High school or less",
  female = "Can you please tell me your gender? Female",
  male = "Can you please tell me your gender? Male",
  non_binary_other = "Can you please tell me your gender? Nonbinary or other",
  caucasian = "For statistical purposes only, can you please tell me your ethnicity? White or Caucasian",
  young_age_18_29 = "What is your age range? 18-29 years",
  young_age_30_39 = "What is your age range? 30-39 years",
  young_age_40_49 = "What is your age range? 40-49 years",
  older_age_50_59 = "What is your age range? 50-59 years",
  older_age_60_69 = "What is your age range? 60-69 years",
  older_age_70_more = "What is your age range? 70 or more years"
)

@@

```{r}
#| label: fig-predictors-pct-summary
#| tbl-cap: "Latest Poll Points by Party"
#| echo: false
#| warning: false
#| message: false

poll_summary <- poll_data %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE),
    highly_educated_pct = mean(`What is the highest level of education you have attained? College graduate`, na.rm = TRUE),
    lower_educated_pct = mean(`What is the highest level of education you have attained? High school or less`, na.rm = TRUE),
    female_pct = mean(`Can you please tell me your gender? Female`, na.rm = TRUE),
    male_pct = mean(`Can you please tell me your gender? Male`, na.rm = TRUE),
    non_binary_other_pct = mean(`Can you please tell me your gender? Nonbinary or other`, na.rm = TRUE),
    caucasian_pct = mean(`For statistical purposes only, can you please tell me your ethnicity? White or Caucasian`, na.rm = TRUE),
    minority_pct = 100 - caucasian_pct,
    age_young_pct = mean(`What is your age range? 18-29 years`, na.rm = TRUE) + 
                    mean(`What is your age range? 30-39 years`, na.rm = TRUE) + 
                    mean(`What is your age range? 40-49 years`, na.rm = TRUE),
    older_pct = mean(`What is your age range? 50-59 years`, na.rm = TRUE) + 
                mean(`What is your age range? 60-69 years`, na.rm = TRUE) + 
                mean(`What is your age range? 70 or more years`, na.rm = TRUE)
  ) %>%
  mutate(across(everything(), ~ replace_na(., 0))) %>%  # Replace NA values with 0
  mutate(across(where(is.numeric), ~ round(., 1)))  # Round to nearest tenth

colnames(poll_summary) <- c("State", "Democrat Pct", "Republican Pct", 
                            "High Educated", "Low Educated", "Female", 
                            "Male", "Nonbinary", "Caucasian", 
                            "Minority Ethnicity", "Young Age", "Older Age")
poll_summary_party <- poll_summary %>%
  select(State, `Democrat Pct`, `Republican Pct`)

poll_summary_demographics <- poll_summary %>%
  select(State, `High Educated`, `Low Educated`, `Female`, 
         `Male`, `Nonbinary`, `Caucasian`, 
         `Minority Ethnicity`, `Young Age`, `Older Age`)

kable(poll_summary_party) %>%
  kable_styling(font_size = 10, position = "center", full_width = FALSE)

# Second table for demographic variables
kable(poll_summary_demographics) %>%
  kable_styling(font_size = 6, position = "center", full_width = FALSE)

```
\newpage  
## Measurement
For this analysis, we selected Emerson as our primary pollster. The rationale behind this choice is explained in detail in @sec-appenA. Emerson’s polls provided essential details, including polling date ranges, party affiliations, sample sizes, and the percentage breakdown of support for different political parties.

We focused on analyzing voter preferences based on party affiliation rather than individual candidate support, as this was more relevant to predicting overall trends. By capturing the broader party dynamics, our model aims to forecast how states are likely to swing in future elections, especially with respect to swing states. This focus will be elaborated in our results section, @sec-results.

Additionally, the Emerson dataset included critical demographic data, such as respondents' political preferences, motivations for voting in the 2024 election, opinions on Donald Trump’s legal battles, and perspectives on major political figures and issues. These details, coupled with information on party affiliation, sources of political news, and other personal attributes, allowed us to paint a richer picture of the voting landscape.

To ensure uniformity across all states, we concentrated on questions that were consistently asked across the entire dataset. This enabled us to perform state-level aggregation and analysis without being limited by state-specific polling questions, ensuring a standardized approach across the board.

## Outcome variables
The initial analysis of the dataset revealed some key trends in party preferences across different states. In @fig-party-results, we illustrate which states exhibit the strongest leanings toward either the Democratic or Republican party. This visualization plays a crucial role in identifying the pivotal swing states that lie close to the center of the graph. These swing states, which show near-equal support for both parties, are of particular interest for future election predictions, as they are likely to determine the overall election outcome.

```{r}
#| label: fig-party-results
#| fig-cap: Example Predictor Variable Visualization
#| echo: false

poll_summary <- poll_data %>%
  group_by(state) %>%   
  summarise(
    dem_total_pct = sum(democrat_pct, na.rm = TRUE),
    rep_total_pct = sum(republican_pct, na.rm = TRUE)
  ) %>%
   mutate(party_diff = dem_total_pct - rep_total_pct)

# Create the vertical bar graph
ggplot(poll_summary, aes(x = party_diff, y = reorder(state, party_diff))) +
  geom_col(aes(fill = party_diff > 0), show.legend = FALSE) +  
  scale_fill_manual(values = c("red", "blue")) +  
  labs(x = "Party Lean (Democrat vs. Republican)", y = "State",
       title = "Party Leaning by State (Democrat vs. Republican)") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 6))

@@

By visualizing the difference between Democratic and Republican preferences across states, we are able to pinpoint which states are firmly partisan and which fall closer to the center, indicating their potential to swing either way in future elections.

## Predictor variables

Next, we take a deeper dive into the various demographic and social factors that influence voter preferences. The predictors we consider include education levels, gender, ethnicity, and age group. These variables are crucial as they shape political preferences and are likely to influence which way a state leans politically.

In @fig-predictors-party-pct, we explore how these factors correlate with party lean across the U.S. states. For instance, higher levels of education may correlate with a greater likelihood of supporting the Democratic party, while other demographic factors such as ethnicity and age may play different roles in shaping political affiliation. This analysis helps us to identify which factors have the greatest influence in each state, offering a clearer understanding of what drives the electorate.

```{r}
#| label: fig-predictors-party-pct
#| fig-cap: Example Predictor Variable Visualization
#| echo: false

# Calculate poll_summary with rounding to the nearest tenth
poll_summary <- poll_data %>%
  group_by(state) %>%
  summarise(
    dem_total_pct = round(sum(.data[[column_refs$democrat]], na.rm = TRUE), 1),
    rep_total_pct = round(sum(.data[[column_refs$republican]], na.rm = TRUE), 1),
    highly_educated_pct = round(mean(.data[[column_refs$highly_educated]], na.rm = TRUE), 1),
    lower_educated_pct = round(mean(.data[[column_refs$lower_educated]], na.rm = TRUE), 1),
    female_pct = round(mean(.data[[column_refs$female]], na.rm = TRUE), 1),
    male_pct = round(mean(.data[[column_refs$male]], na.rm = TRUE), 1),
    non_binary_other_pct = round(mean(.data[[column_refs$non_binary_other]], na.rm = TRUE), 1),
    caucasian_pct = round(mean(.data[[column_refs$caucasian]], na.rm = TRUE), 1),
    minority_pct = round(100 - caucasian_pct, 1),
    age_young_pct = round(mean(.data[[column_refs$young_age_18_29]], na.rm = TRUE) + 
                          mean(.data[[column_refs$young_age_30_39]], na.rm = TRUE) + 
                          mean(.data[[column_refs$young_age_40_49]], na.rm = TRUE), 1),
    older_pct = round(mean(.data[[column_refs$older_age_50_59]], na.rm = TRUE) + 
                      mean(.data[[column_refs$older_age_60_69]], na.rm = TRUE) + 
                      mean(.data[[column_refs$older_age_70_more]], na.rm = TRUE), 1)
  ) %>%
  mutate(party_diff = round(dem_total_pct - rep_total_pct, 1)) %>%
  drop_na()

# Select relevant numeric columns for correlation matrix, excluding PartyDifference
corr_data <- poll_summary %>%
  select(
    Democrat = dem_total_pct,
    Republican = rep_total_pct,
    HighlyEducated = highly_educated_pct,
    LowerEducated = lower_educated_pct,
    Female = female_pct,
    Male = male_pct,
    NonBinary = non_binary_other_pct,
    Caucasian = caucasian_pct,
    Minority = minority_pct,
    YoungAge = age_young_pct,
    OlderAge = older_pct
  )

# Calculate the correlation matrix
cor_matrix <- cor(corr_data, use = "complete.obs")

# Set correlations less than -0.9 to NA
cor_matrix[cor_matrix < -0.9] <- NA

# Visualize the correlation matrix without hierarchical clustering
ggcorrplot(cor_matrix, 
           hc.order = FALSE,  
           type = "lower", 
           lab = TRUE, 
           lab_size = 2,  # Size for correlation labels
           title = "Correlation Matrix of Predictors and Party Lean",
           colors = c("blue", "white", "red"),  
           legend.title = "Correlation") +
  theme(
    plot.title = element_text(size = 12, hjust = 0.5),  # Title size and centered
    axis.text = element_text(size = 1),  # Axis text size
    legend.text = element_text(size = 10),  # Legend text size
    legend.title = element_text(size = 10),  # Legend title size
    axis.title = element_text(size = 1),  # Axis labels size
  )

```
The correlation matrix in @fig-predictors-party-pct provides a detailed look at how various predictors are related to party preferences. From this, we can see which predictors—such as education or gender—are more strongly associated with Democratic or Republican leanings. The blue shading indicates stronger correlations with Democratic support, while red represents stronger correlations with Republican support.

This analysis offers valuable insights into how specific demographic factors shape the political landscape in different states, allowing us to better predict electoral outcomes based on shifting demographic trends.

# Pollster Methodology Overview (Appendix A) {#sec-appenA}  
Append a detailed review of a chosen pollster's methodology, survey techniques, strengths, and weaknesses.

@@

# Results {#sec-results}
